---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# BayesGrowth

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/jonathansmart/BayesGrowth.svg?branch=master)](https://travis-ci.org/jonathansmart/BayesGrowth)
<!-- badges: end -->

BayesGrowth combines length-at-age modelling for fisheries with MCMC implemented using JAGS and the [rjags](https://cran.r-project.org/web/packages/rjags/index.html) package. Growth modelling using models such as the von Bertalanffy growth model involves three parameters: $L_{\infty}$, $k$ and either $L_{0}$ or $t_{0}$. Two of these parameters: $L_{0}$ and $L_{\infty}$ have direct biological meaning as the size-at-birth and maximum length, respectively. This package provides the tools to run an MCMC model with these two parameters treated as size-at-birth and maximum length using a JAGS model. This MCMC model is pre-specified and built into wrapper functions.

The user can therefore run an MCMC growth model using knowledge of species length-at-birth and maximum size ast priors.

## Installation

You can install the released version of BayesGrowth from [Github](https://github.com/jonathansmart/BayesGrowth) with:

``` {r, eval = FALSE}
devtools::install_github("jonathansmart/BayesGrowth")
```

## Usage

The main `BayesGrowth` function is `Estimate_MCMC_Growth` which is the wrapper function around an rjags model. It requires a data input that includes Age and Length, a model needs to be specified (several options are available) and the priors must be specified. Priors include the max size with an error, length-at-birth with an error, an upper value for $k$ and $\sigma$. These latter two parameters have no informative priors so just need senscible upper bounds. Many fish species (including this example) have a size at birth of zero. Therefore, this can be used along with a very small error to indicate that this is a certain value. The `L0.se` argument cannot be zero, although the model is specified to truncate it to zero and keep growth positive.

```{r model,message=FALSE, warning=FALSE}
library(BayesGrowth)

data("example_data")
## Biological info - lengths in mm
max_size <- 440
max_size_se <- 4.4
birth_size <- 0
birth_size_se <- 0.001 # an se cannot be zero

# Use the function to estimate the JAGS model
results <- Estimate_MCMC_Growth(example_data, Model = "VB" ,
                     iter = 10000,
                     Linf = 440,
                     Linf.se = 4.4,
                     L0 = 0,
                     sigma.max = 100,
                     L0.se = .001,
                     k.max = 1)

```


The function returns the rjags outputs which is an object of class 'mcmc.list'
```{r output,message=FALSE}
head(results)
```

Therefore, all of the diagnostics from the rjags library can be used.
```{r Diagnostics,message=FALSE,  fig.height = 10, fig.width = 8}
summary(results)
plot(results,density = T, smooth = F)
```

Additional `BayesGrowth` functions are available that help the user manipulate the objects. The 
`Calculate_MCMC_growth_curve` function will provide confidence intervals around the growth curve based on MCMC percentiles. This is essentially a wrapper around the `tidybayes::mean_qi()` function which means it can be passed straight into a ggplot with the `tidybayes::geom_line_ribbon` function.

```{r plot,message=FALSE, fig.height = 6, fig.width = 8, warning=FALSE}

growth_curve <- Calculate_MCMC_growth_curve(results, Model = "VB",
                                            max.age = max(example_data$Age), probs = .95)
library(tidybayes)
library(ggplot2)

ggplot(growth_curve, aes(Age, LAA))+
  geom_point(data = example_data, aes(Age, Length), alpha = .3)+
  geom_lineribbon(size = 1.5, fill = "steelblue") +
  labs(y = "Total Length (mm)", x = "Age (yrs)")+
  scale_y_continuous(expand = c(0,0))+
  scale_x_continuous(expand = c(0,0), breaks = seq(0,13,1))+
  theme_bw()+
  theme(text = element_text(size = 14))

```

This represents a much improved fit over a standard non-linear estimated model, even if the length-at-birth were fixed at zero. Here the fit is compared using an nls model fit using the [AquaticLifeHistory package](https://github.com/jonathansmart/AquaticLifeHistory).


```{r compare plot,message=FALSE, fig.height = 6, fig.width = 8, warning=FALSE, echo=FALSE}
 
library(AquaticLifeHistory)
fixed_mod <- Estimate_Growth(example_data,n.bootstraps = 1000, Birth.Len = 0, models = "VB", plots = FALSE)
free_mod <- Estimate_Growth(example_data,n.bootstraps = 1000, models = "VB", plots = FALSE)

ggplot(growth_curve, aes(Age, LAA))+
  geom_line(data = free_mod$Estimates,inherit.aes = FALSE,
            aes(Age, AVG, col = "nls model - free L0"), size = 1.5)+

  geom_line(data = fixed_mod$Estimates,inherit.aes = FALSE,
            aes(Age, AVG, col = "nls model - fixed L0"), size = 1.5)+
  geom_point(data = example_data, aes(Age, Length), alpha = .3)+
  geom_lineribbon( aes(fill = "MCMC model", col = "MCMC model"),size = 1.5, alpha = .5) +
   geom_line(aes(Age, LAA, col = "MCMC model"), size = 1.5)+
  labs(y = "Total Length (mm)", x = "Age (yrs)")+
  scale_color_viridis_d(name = "Model", begin = .3, end = .8)+
  scale_fill_viridis_d(begin = .5, end = .6)+
  guides(fill = FALSE)+
  scale_y_continuous(expand = c(0,0))+
  scale_x_continuous(expand = c(0,0), breaks = seq(0,13,1))+
  theme_bw()+
  theme(text = element_text(size = 14))

```

